Multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-08-09 paper/mcd2.tex
edits start of section 6

The theory presented in \cite{Tai2003} is extended as follows:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective---examples are explored in Sections \ref{sec:vi} and \ref{sec:results}---but we prove convergence in norm at the same rate as the original method converges in energy (Section \ref{sec:convergence}). % THIS IS THE HOPE
\end{enumerate}
Furthermore the multilevel CD algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} are extended in three practical ways:
\begin{enumerate}
\setcounter{enumi}{1} % resume at (ii)
\item We make finite element (FE) implementations of multilevel CD algorithm more effective for nonlinear problems by applying a full approximation storage (FAS; see \cite{Brandt1977,Bruneetal2015}) approach.  Results from the resulting \emph{nonlinear multilevel constraint decomposition} (NMCD) algorithm, which avoids global (Newton) linearization (compare \cite{GraeserKornhuber2009}), are shown in Section \ref{sec:results} for several nonlinear problems.
\item Our implementation permits ``box'' constraints, namely upper and lower obstacles.
\item In Sections \ref{sec:fe} and \ref{sec:multilevel} we explain why, given the manner in which constraint sets are decomposed in the multilevel context \cite{GraeserKornhuber2009}, multilevel ``up-smoothing'' is more efficient than ``down-smoothing''.  This observation seems to be new; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  A strong preference for up-smoothing is special to multilevel CD methods and does not arise in unconstrained problems.
\end{enumerate}

@@

Let $\cK \subset \cV$ be a nonempty closed and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and a linear \emph{source functional} $\ell\in \cV'$ we consider the following \emph{variational inequality} (VI) for the (exact) solution $u^*\in \cK$, if it exists:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, the source term $\ell$ is not strictly needed, and indeed by redefining $f$ we may take $\ell=0$, but its presence clarifies algorithms in Section \ref{sec:multilevel}.

@@

In Section \ref{sec:multilevel} we will introduce discretizations, and practical algorithms for finite-dimensional VI problems, but observe here that this CD concept applies at the level of the continuum problem.  The following two obstacle problem examples illustrate this.  The first is an overlapping domain decomposition.

@@

Our second example is a disjoint frequency decomposition.  A multilevel CD, e.g.~the one proposed in Sections \ref{sec:fe} and \ref{sec:multilevel}, approximates such a frequency decomposition.

@@

When $f$ has all the nice properties listed in Example \ref{ex:fnice} then a solver can take advantage of implementation efficiencies which are unavailable in general.  The FE approximation of VI \eqref{eq:vi}---see Section \ref{sec:fe}---can be solved in any incremental and efficient manner following the techniques in \cite{GraeserKornhuber2009,Tai2003}, namely by computations over hat function supports.  The actual solution process for VI \eqref{eq:linearlocalvi} is not really the concern here; we are observing that the storage needed for problem \eqref{eq:linearlocalvi}, and the cost of residual evaluation, are smaller than in the general case.  For more difficult problems certain additional ideas, illustrated in Section \ref{sec:multilevel}, are needed for practical implementation.  Specifically, we will extend a multilevel CD iteration to a method for nonlinear functions $f:\cK\to\cV'$, i.e.~defined only on the constraint set $\cK$, by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.

@@

We make the following observations:
\begin{itemize}
\item The constants $C_1,C_2$ generally depend on $m$, the number of sets in the CD.  In a multilevel CD method (Section \ref{sec:multilevel}) this number is logarithmic in the number of degrees of freedom (nodes) in the fine-level mesh.
\item Regarding the convergence proof below, if the iterates from Algorithm \ref{alg:basiccd-add} are bounded in $\mathcal{K}$ then constants $C_1,C_2$ are permitted to depend continuously on $u,v,w_{ij},v_i,y_i$ \cite{Tai2003}.
\item Bound \eqref{as:lipschitzrestrictions} addresses only the CD, and not the nonlinear operator $f$.  Tai \cite{Tai2003} gives a value of $C_1$ for certain obstacle problem CDs using $P_1$ FE spaces over shape-regular and quasi-uniform triangulations.  In particular, $C_1$ is known for the standard multilevel hierarchy, so the cases demonstrated in Section \ref{sec:results} are covered.
\item If $f$ is defined on all of $\mathcal{V}$ then one may verify \eqref{as:lipschitzresidual} over a subspace, versus constraint, decomposition.  For example, \eqref{as:lipschitzresidual} is verified for the classical obstacle problem over arbitrary $w_{ij} \in \cV$.  (Unconstrained versions of \eqref{as:lipschitzrestrictions}, \eqref{as:lipschitzresidual} are (13), (14) in \cite{TaiXu2002}, respectively.)
\item If $f$ is Lipschitz then the existence of $C_2$ is clear \cite{TaiXu2002}.  However, our convergence rate bound is improved by taking the smallest $C_2$ for which \eqref{as:lipschitzresidual} holds.
\end{itemize}

@@

\section{Multilevel finite elements} \label{sec:fe}

We will approximately solve VI \eqref{eq:vi} for certain nonlinear obstacle problems, over two-dimensional computational domains $\Omega \subset \RR^2$ and Banach space $\mathcal{V}=W_0^{1,p}(\Omega)$, using piecewise-linear finite element (FE) spaces over triangular meshes.  The targeted mesh will be the finest level in a hierarchy of meshes constructed by refinement of a coarsest mesh.

@@

Let $\tilde{\mathcal{V}}^j$ be the continuous, piecewise-polynomial FE space $P_1$,\footnote{The FE method herein apparently extends to $Q_1$ spaces over quadrilateral meshes \cite{Elmanetal2014}.  However, for $k>1$ the $P_k$ space seems not to be suitable because the monotone restriction operator, \eqref{eq:fe:monotonerestriction} below, depends on element-wise monotonicity.  That is, for $P_1$ or $Q_1$ if the nodal values of an FE function (non-strictly) increase then the FE function also increases.  As shown by Figure 1.7 of \cite{Elmanetal2014}, for example, such monotonicity is not true for $P_2$.} over $\mathcal{T}^j$, with arbitrary boundary values.  Let $\mathcal{V}^j \subset \tilde{\mathcal{V}}^j$ be the subspace with zero boundary values.  For any $p\ge 1$, $\tilde{\mathcal{V}}^j$ is a subspace of $W^{1,p}(\Omega)$ and $\mathcal{V}^j$ is a subspace of $\mathcal{V}=W_0^{1,p}(\Omega)$ \cite{Elmanetal2014}.  Our choices imply the following nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}.  \label{eq:fe:nestedspaces}
\end{equation}

Our only need for the arbitrary-boundary-value spaces $\tilde{\mathcal{V}}^j$ is for the constraint.  Suppose $\gamma^J \in \tilde{\mathcal{V}}^j$ denotes the \emph{obstacle}, a function satisfying $\gamma^J \le 0$ at boundary nodes of $\mathcal{T}^J$, and define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,v \ge \gamma^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions.  Note that admissible functions from $\mathcal{K}^J$ have zero boundary values.

@@

\section{Nonlinear multilevel constraint decomposition} \label{sec:multilevel}

>>>>

The theory presented in \cite{Tai2003} is extended as follows:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective---examples are explored in Sections \ref{sec:vi} and \ref{sec:results}---but we prove convergence in norm at the same rate as the original method converges in energy (Section \ref{sec:convergence}). % THIS IS THE HOPE
\end{enumerate}
Furthermore the multilevel CD algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} are extended in three practical ways:
\begin{enumerate}
\setcounter{enumi}{1} % resume at (ii)
\item We make finite element (FE) implementations of multilevel CD algorithm more effective for nonlinear problems by applying a full approximation storage (FAS; see \cite{Brandt1977,Bruneetal2015}) approach.  Results from the resulting \emph{nonlinear multilevel constraint decomposition} (NMCD) algorithm, which avoids global (Newton) linearization (compare \cite{GraeserKornhuber2009}), are shown in Section \ref{sec:results} for several nonlinear problems.
\item Our implementation permits ``box'' constraints, namely upper and lower obstacles.
\item In Sections \ref{sec:multilevel} and \ref{sec:vcycle} we explain why, given the manner in which constraint sets are decomposed in the multilevel context \cite{GraeserKornhuber2009}, multilevel ``up-smoothing'' is more efficient than ``down-smoothing''.  This observation seems to be new; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  A strong preference for up-smoothing is special to multilevel CD methods and does not arise in unconstrained problems.
\end{enumerate}

@@

Let $\cK \subset \cV$ be a nonempty closed and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and a linear \emph{source functional} $\ell\in \cV'$ we consider the following \emph{variational inequality} (VI) for the (exact) solution $u^*\in \cK$, if it exists:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, the source term $\ell$ is not strictly needed, and indeed by redefining $f$ we may take $\ell=0$, but its presence clarifies the algorithm in Section \ref{sec:vcycle}.

@@

In Sections \ref{sec:multilevel} and \ref{sec:vcycle} we will introduce discretizations and a practical algorithms for finite-dimensional VI problems, but observe here that this CD concept applies at the level of the continuum problem.  The following two obstacle problem examples illustrate this.  The first is an overlapping domain decomposition.

@@

Our second example is a disjoint frequency decomposition.  A multilevel CD, e.g.~the one proposed in Section \ref{sec:multilevel}, approximates such a frequency decomposition.

@@

When $f$ has all the nice properties listed in Example \ref{ex:fnice} then a solver can take advantage of implementation efficiencies which are unavailable in general.  The FE approximation of VI \eqref{eq:vi}---see Sections \ref{sec:multilevel} and \ref{sec:vcycle}---can be solved in any incremental and efficient manner following the techniques in \cite{GraeserKornhuber2009,Tai2003}, namely by computations over hat function supports.  The actual solution process for VI \eqref{eq:linearlocalvi} is not really the concern here; we are observing that the storage needed for problem \eqref{eq:linearlocalvi}, and the cost of residual evaluation, are smaller than in the general case.  For more difficult problems certain additional ideas, illustrated in Section \ref{sec:vcycle}, are needed for practical implementation.  Specifically, we will extend a multilevel CD iteration to a method for nonlinear functions $f:\cK\to\cV'$, i.e.~defined only on the constraint set $\cK$, by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.

@@

We make the following observations:
\begin{itemize}
\item The constants $C_1,C_2$ generally depend on $m$, the number of sets in the CD.  In a multilevel CD method (Sections \ref{sec:multilevel} and \ref{sec:vcycle}) this number is logarithmic in the number of degrees of freedom (nodes) in the fine-level mesh.
\item Regarding the convergence proof below, if the iterates from Algorithm \ref{alg:basiccd-add} are bounded in $\mathcal{K}$ then constants $C_1,C_2$ are permitted to depend continuously on $u,v,w_{ij},v_i,y_i$ \cite{Tai2003}.
\item Bound \eqref{as:lipschitzrestrictions} addresses only the CD, and not the nonlinear operator $f$.  Tai \cite{Tai2003} gives a value of $C_1$ for certain obstacle problem CDs using $P_1$ FE spaces over shape-regular and quasi-uniform triangulations.  In particular, $C_1$ is known for the standard multilevel hierarchy, so the cases demonstrated in Section \ref{sec:results} are covered.
\item If $f$ is defined on all of $\mathcal{V}$ then one may verify \eqref{as:lipschitzresidual} over a subspace, versus constraint, decomposition.  For example, \eqref{as:lipschitzresidual} is verified for the classical obstacle problem over arbitrary $w_{ij} \in \cV$.  (Unconstrained versions of \eqref{as:lipschitzrestrictions}, \eqref{as:lipschitzresidual} are (13), (14) in \cite{TaiXu2002}, respectively.)
\item If $f$ is Lipschitz then the existence of $C_2$ is clear \cite{TaiXu2002}.  However, our convergence rate bound is improved by taking the smallest $C_2$ for which \eqref{as:lipschitzresidual} holds.
\end{itemize}

@@

\section{Multilevel finite element constraint decompositions} \label{sec:multilevel}

We will approximately solve VI \eqref{eq:vi} for certain nonlinear obstacle problems, over two-dimensional computational domains $\Omega \subset \RR^2$ and the Banach space $\mathcal{V}=W_0^{1,p}(\Omega)$, using piecewise-linear finite element (FE) spaces over triangular meshes.  The targeted mesh will be the finest level in a hierarchy of meshes constructed by refinement of a coarsest mesh.

@@

Let $\mathcal{V}^j$ be the continuous, piecewise-linear FE space $P_1$ over $\mathcal{T}^j$ with zero boundary values.  Note $\mathcal{V}^j$ is a subspace of $\mathcal{V}=W_0^{1,p}(\Omega)$ \cite{Elmanetal2014}, and observe the following nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}.  \label{eq:fe:nestedspaces}
\end{equation}

Let $\tilde{\mathcal{V}}^j$ denote the set of functions on the nodes of $\mathcal{T}^j$ which take their values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  Our purpose for such extended-value functions is to define ``box constraints'' \cite{BensonMunson2006} as follows.  At the finest-level, suppose $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad \overline{\gamma}^J > -\infty, \qquad \text{and} \qquad \underline{\gamma}^J|_{\partial\Omega} \le 0 \le \overline{\gamma}^J|_{\partial\Omega}. \label{eq:fe:boxconstraintrequirements}
\end{equation}
We call $\underline{\gamma}^J$ the \emph{lower obstacle} and $\overline{\gamma}^J$ the \emph{upper obstacle}.  Where $\underline{\gamma}^J=-\infty$ there is no lower obstacle and where $\overline{\gamma}^J=+\infty$ there is no upper obstacle.  Now define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,\underline{\gamma}^J \le v \le \overline{\gamma}^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions.  Note that admissible functions from $\mathcal{K}^J$ have zero boundary values and that elements of $\mathcal{K}^J$ never, themselves, take on values $\pm\infty$.

The FE method herein apparently extends to $Q_1$ spaces over quadrilateral meshes \cite{Elmanetal2014}.  However, for $k>1$ the $P_k$ space seems not to be suitable because the monotone restriction operator, \eqref{eq:fe:monotonerestriction} below, depends on element-wise monotonicity.  That is, for $P_1$ or $Q_1$ if the nodal values of an FE function (non-strictly) increase then the FE function also increases.  As shown by Figure 1.7 of \cite{Elmanetal2014}, for example, such monotonicity is not true for $P_2$.

@@

\section{The nonlinear multilevel constraint decomposition V-cycle} \label{sec:vcycle}
